---
title: "Project"
author: "Team 24"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r load-packages, message = FALSE, include = FALSE}
library(knitr)
library(tidyverse)
library(broom)
library(htmltools)
library(scales)
library(grid)
library(tinytex)
library(gridExtra)
library(knitr)
library(tidyverse)
library(broom)
library(htmltools)
library(scales)
library(grid)
library(tinytex)
library(gridExtra)
```

laoding data to a DF

```{r, warning=FALSE}
data <- read.csv("data.csv")
```

droping NONE values from the DF that represent missing values.
ranging from 0 to the highest degree of that feature. 
```{r warning=FALSE}

data <- data[data$EDUC != -9, ]

data <- data[data$EMPLOY != -9, ]

data <- data[data$LIVARAG != -9, ]

data <- data[data$PRIMINC != -9, ]

data$EDUC <- data$EDUC - 1

data$PRIMINC <- 5 - data$PRIMINC

data$EMPLOY <- 4 - data$EMPLOY

data$LIVARAG <- data$LIVARAG - 1

```

normalizing all the rows within a range of 0-1 and creating a new column
```{r, warning=FALSE}
features <- c("EDUC", "EMPLOY", "LIVARAG", "PRIMINC")

for (col in features) {
  data[[col]] <- (data[[col]] - min(data[[col]], na.rm = TRUE)) / (max(data[[col]], na.rm = TRUE) - min(data[[col]], na.rm = TRUE))
}

data$SocialScore <- rowSums(data[, c("EDUC", "EMPLOY", "LIVARAG", "PRIMINC")], na.rm = TRUE)

```

making a hisplot of the Social Score to determine the place to split them

```{r, warning=FALSE}
hist(data$SocialScore, breaks = 25, col = "blue", xlab = "Social Score", ylab = "Frequency", main = "Histogram of Social Score")
abline(v = 2.5, col = "red")
```

spliting the DF into tow categories and droping missing values from NOPRIOR category

```{r, warning=FALSE}
data <- data[data$NOPRIOR != 5 & data$NOPRIOR != -9, ]

social_H <- data[data$SocialScore >= 2.5, ]
social_L <- data[data$SocialScore < 2.5, ]

```


filter high and low to the needed features

```{r, warning=FALSE}

filtered_df_low <- social_L
filtered_df_low <- select(filtered_df_low, EDUC, EMPLOY , LIVARAG, PRIMINC,SocialScore,SUB1,NOPRIOR)
```


```{r, warning=FALSE}
filtered_df_high <- social_H
filtered_df_high <- select(filtered_df_high, EDUC, EMPLOY , LIVARAG, PRIMINC,SocialScore ,SUB1,NOPRIOR)
```


divition


alcohol:
```{r, warning=FALSE}

alcohol_high <- subset(filtered_df_high, SUB1 =="2")
alcohol_low <- subset(filtered_df_low, SUB1 =="2")

alcohol_nopriorhigh <- alcohol_high$NOPRIOR
alcohol_nopriorlow <- alcohol_low$NOPRIOR
t_test_result <- t.test(alcohol_nopriorhigh, alcohol_nopriorlow)

t_test_result
```


Cocaine and crack:
```{r, warning=FALSE}

Cocaine_crack_high <- subset(filtered_df_high, SUB1 =="3"& NOPRIOR != -9)
Cocaine_crack_low <- subset(filtered_df_low, SUB1 =="3"& NOPRIOR != -9)

Cocaine_cracknopriorhigh <- Cocaine_crack_high$NOPRIOR
Cocaine_cracknopriorlow <- Cocaine_crack_low$NOPRIOR
t_test_result <- t.test(Cocaine_cracknopriorhigh, Cocaine_cracknopriorlow)

t_test_result
```


Marijuana:
```{r, warning=FALSE}
Marijuana_high <- subset(filtered_df_high, SUB1 =="4" & NOPRIOR != -9)
Marijuana_low <- subset(filtered_df_low, SUB1 =="4" & NOPRIOR != -9)

Marijuana_nopriorhigh <- Marijuana_high$NOPRIOR
Marijuana_nopriorlow <- Marijuana_low$NOPRIOR
t_test_result <- t.test(Marijuana_nopriorhigh, Marijuana_nopriorlow)

t_test_result
```


Heroin:
```{r, warning=FALSE}
Heroin_high <- subset(filtered_df_high, SUB1 =="5" & NOPRIOR != -9)
Heroin_low <- subset(filtered_df_low, SUB1 =="5" & NOPRIOR != -9)

Heroin_nopriorhigh <- Heroin_high$NOPRIOR
Heroin_nopriorlow <- Heroin_low$NOPRIOR
t_test_result <- t.test(Heroin_nopriorhigh, Heroin_nopriorlow)

t_test_result

```


opiates:
```{r, warning=FALSE}
opiates_high <- subset(filtered_df_high, SUB1 =="7" & NOPRIOR != -9)
opiates_low <- subset(filtered_df_low, SUB1 =="7" & NOPRIOR != -9)

opiates_nopriorhigh <- opiates_high$NOPRIOR
opiates_nopriorlow <- opiates_low$NOPRIOR
t_test_result <- t.test(opiates_nopriorhigh, opiates_nopriorlow)

t_test_result

```

speed:
```{r, warning=FALSE}

speed_high <- subset(filtered_df_high, SUB1 =="10" & NOPRIOR != -9)
speed_low <- subset(filtered_df_low, SUB1 =="10" & NOPRIOR != -9)

speed_nopriorhigh <- speed_high$NOPRIOR
speed_nopriorlow <- speed_low$NOPRIOR
t_test_result <- t.test(speed_nopriorhigh, speed_nopriorlow)

t_test_result

```


Benzodiazepines:
```{r, warning=FALSE}
Benzodiazepines_high <- subset(filtered_df_high, SUB1 =="13" & NOPRIOR != -9)
Benzodiazepines_low <- subset(filtered_df_low, SUB1 =="13" & NOPRIOR != -9)

Benzodiazepines_nopriorhigh <- Benzodiazepines_high$NOPRIOR
Benzodiazepines_nopriorlow <- Benzodiazepines_low$NOPRIOR
t_test_result <- t.test(Benzodiazepines_nopriorhigh, Benzodiazepines_nopriorlow)

t_test_result
numRows3 <- nrow(Benzodiazepines_low)
print(numRows3)

```






```{r ,fig.height=4, fig.width = 10}
# Calculate the mean values for each pair
mean_values <- c(mean(alcohol_nopriorhigh), mean(alcohol_nopriorlow),
                 mean(Benzodiazepines_nopriorhigh), mean(Benzodiazepines_nopriorlow),
                 mean(Cocaine_cracknopriorhigh), mean(Cocaine_cracknopriorlow),
                 mean(Heroin_nopriorhigh), mean(Heroin_nopriorlow),
                 mean(Marijuana_nopriorhigh), mean(Marijuana_nopriorlow),
                 mean(opiates_nopriorhigh), mean(opiates_nopriorlow),
                 mean(speed_nopriorhigh), mean(speed_nopriorlow))

# Create a bar plot
colors <- c("brown3", "brown3", "orange1", "orange1", "yellow", "yellow", "forestgreen", "forestgreen", "darkturquoise", "darkturquoise", "mediumblue", "mediumblue", "mediumorchid", "mediumorchid")
y_min <- 0   # Minimum value
y_max <- 2  # Maximum value

# Set graphical parameter xpd to TRUE
par(xpd = TRUE)

# Create the bar plot
bp <- barplot(mean_values, beside = TRUE,
        xlab = "", ylab = "Mean Value",
        main = "Comparison of Mean Values", col = colors, ylim = c(y_min, y_max))

# Rotate x-axis labels by 45 degrees using text() function
text(bp, par("usr")[3] - 0.1, srt = 45, adj = 1, xpd = TRUE,
     labels = c("alcohol_high","alcohol_low", "Benzo_high","Benzo_low" , "Cocaine_high", "Cocaine_low" , "Heroin_high","Heroin_low", "Marijuana_high", "Marijuana_low", "opiates_high","opiates_low" ,"speed_high" , "speed_low"))


```



-------------------------------------------------------
# Read the drugs data from a CSV file into a data frame
```{r}
drugs = read.csv("data.csv", header = TRUE, sep = ",", dec = ".",)
drugs
```
# Filter out rows with missing values for specific columns
```{r}
drugs_data <- drugs[drugs$NOPRIOR != -9, ]
drugs_data <- drugs_data[drugs_data$EDUC != -9, ]
drugs_data <- drugs_data[drugs_data$PRIMINC != -9, ]
drugs_data <- drugs_data[drugs_data$LIVARAG != -9, ]
drugs_data <- drugs_data[drugs_data$EMPLOY != -9, ]
drugs_data <- drugs_data[drugs_data$SUB1 != -9, ]

drugs_data
```
# Convert values 1, 2, 3, 4, 5 to 1 in the NOPRIOR column
```{r, eval=FALSE}
drugs_data_binary <- drugs_data
drugs_data_binary$NOPRIOR <- ifelse(drugs_data_binary$NOPRIOR %in% c(1, 2, 3, 4, 5), 1, drugs_data_binary$NOPRIOR)
```


# Convert selected columns to factor variables
```{r, eval=FALSE}
drugs_data_binary <- drugs_data_binary %>%
  mutate(
    PRIMINC = factor(PRIMINC),
    EMPLOY = factor(EMPLOY),
    DETNLF = factor(LIVARAG),
    SUB1 = factor(SUB1),
    NOPRIOR = factor(NOPRIOR),
  )
```
# Split to train and test
```{r split, , eval=FALSE}
set.seed(1122)
drugs_split <- initial_split(drugs_data_binary)
drugs_train <- training(drugs_split)
drugs_test <- testing(drugs_split)
```

# The recipe
```{r}
drugs_rec <- recipe(NOPRIOR ~ EDUC + PRIMINC + LIVARAG + EMPLOY + SUB1, data = drugs_data_binary) %>%
  step_dummy(all_nominal(), -all_outcomes(), -c(EDUC)) %>%
  step_integer(EDUC) %>%
  step_zv(all_predictors())

```

# The model - logistic reggression
```{r}
drugs_mod <- logistic_reg() %>% 
  set_engine("glm")

```

# Workflow
```{r}
drugs_wflow <- workflow() %>% 
  add_model(drugs_mod) %>% 
  add_recipe(drugs_rec)

```

# Fitting the model
```{r}
drugs_fit <- fit(drugs_wflow, data = drugs_train)
```
```{r}
drugs_fit
```
# Extracting model coefficients
```{r}
tidy(drugs_fit) %>% print(n = 100)
```

# Making predictions
```{r}
drugs_pred <- predict(drugs_fit, drugs_test, type = "prob") %>% 
  bind_cols(drugs_test)
drugs_pred
```

# Performance evaluation and confusion matrix
```{r}
cutoff_prob <- 0.65
confusion_matrix <- drugs_pred %>%
  mutate(
    NOPRIOR      = if_else(NOPRIOR == 1, "1", "0"),
    NOPRIOR_pred = if_else(.pred_1 > cutoff_prob, "PRED 1", "PRED 0")
    ) %>%
  count(NOPRIOR_pred, NOPRIOR) %>%
  pivot_wider(names_from = NOPRIOR, values_from = n)

confusion_matrix
```

```{r}
TP <- confusion_matrix[2, "1"]  # True Positive
FP <- confusion_matrix[2, "0"]  # False Positive
FN <- confusion_matrix[1, "1"]  # False Negative

precision <- TP / (TP + FP)
recall <- TP / (TP + FN)
f1_score <- 2 * (precision * recall) / (precision + recall)

precision <- pull(precision, 1)
recall <- pull(recall, 1)
f1_score <- pull(f1_score, 1)

cat("The precision is", precision, "\n")
cat("The recall is", recall, "\n")
cat("The F1 Score is", f1_score, "\n")
```
# ROC analysis
```{r}
drugs_pred %>%
  roc_auc(
    truth = NOPRIOR,
    .pred_1,
    event_level = "second"
  ) 
```

```{r}
drugs_pred %>%
  roc_curve(
    truth = NOPRIOR,
    .pred_1,
    event_level = "second"
  ) %>%
  autoplot()
```



#  Bar plots to visualize the distribution of NOPRIOR with other variables (EMPLOY, LIVARAG, and SUB1)

```{r}
ggplot(drugs_data, aes(x = EMPLOY, fill = factor(NOPRIOR))) +
  geom_bar(position = "fill") +
  labs(x = "EMPLOY", y = "Proportion", fill = "NOPRIOR") +
  ggtitle("Bar Plot: Distribution of NOPRIOR by EMPLOY")
```
```{r}
ggplot(drugs_data, aes(x = LIVARAG, fill = factor(NOPRIOR))) +
  geom_bar(position = "fill") +
  labs(x = "LIVARAG", y = "Proportion", fill = "NOPRIOR") +
  ggtitle("Bar Plot: Distribution of NOPRIOR by LIVARAG")
```


```{r}
ggplot(drugs_data, aes(x = SUB1, fill = factor(NOPRIOR))) +
  geom_bar(position = "fill") +
  labs(x = "SUB1", y = "Proportion", fill = "NOPRIOR") +
  ggtitle("Bar Plot: Distribution of NOPRIOR by SUB1")
```



### Data README

```{r include_data_readme, comment=''}
cat(readLines('../data/README.md'), sep = '\n')
```

### Source code

```{r, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```
